---
title: "Building a Graph in ggplot2"
editor: source
format:
  closeread-html:
    remove-header-space: true
    code-tools: true
    cr-style:
      narrative-text-color-sidebar: white
      narrative-font-family: 'Georgia, "Times New Roman", Times, serif'
      narrative-font-size: 1.1rem
      narrative-sidebar-width: minmax(625px, 1fr)
      narrative-border-radius: 5px
      narrative-background-color-overlay: "#111111dd"
      narrative-background-color-sidebar: "#111111"
      section-background-color: white
sidebar: false
---

```{=html}
<style>

  #quarto-document-content h1 {
    font-weight: 500;
    font-size: 2.3em;
    line-height: 1.1;
    margin-top: 0;
  }

  .cr-section  .narrative-col  > *:first-child {
    padding-block-start: 10svh;
  }
  
  .cr-section {
    margin-bottom: 30px;
  }
</style>
```

```{r}
#| include: false
library(ggplot2)
library(bcscr)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  fig.align = "center"
)
```

::: {.cr-section}

# Building a Graph in ggplot2

:::{focus-on=""}
Packages that we will use:

```{r}
#| echo: true
#| eval: false
library(ggplot2)
library(bcscr)    # for m111survey data
```
:::

::: {#cr-mygraph-caption .scale-to-fill}
```{r}
#| fig-cap: "Scatterplot of GPA vs. fastest speed driven, for MAT 111 students."
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex)) +
  labs(
    x = "fastest speed ever driven (mph)",
    y = "grade-point average",
    title = "Not much overall relationship between speed and GPA!",
    subtitle = "(the fellers seem to drive faster and have lower GPAs)"
  )
```
:::

Here is our target graph.  Let's build it, step by step. @cr-mygraph-caption




::: {#cr-mygraph .scale-to-fill}
```{r}
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex)) +
  labs(
    x = "fastest speed ever driven (mph)",
    y = "grade-point average",
    title = "Not much overall relationship between speed and GPA!",
    subtitle = "(the fellers seem to drive faster and have lower GPAs)"
  )
```
:::


:::{#cr-frame .scale-to-fill}
```{r}
#! echo: true
#! eval: false
ggplot(m111survey, aes(x = fastest, y = GPA))
```
:::

:::{focus-on="cr-frame"}
First, set up the frame.

```r
ggplot(m111survey, aes(x = fastest, y = GPA))
```
:::



:::{focus-on="cr-glyphs"}
Next, the glyphs.

```r
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point()
```
:::

:::{#cr-glyphs .scale-to-fill}
```{r}
#! echo: true
#! eval: false
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point()
```
:::



:::{focus-on="cr-aes"}
You can include ane or more aesthetic mappings for the glyph.

```r
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex))
```
:::

:::{#cr-aes .scale-to-fill}
```{r}
#! echo: true
#! eval: false
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex))
```
:::



:::{focus-on="cr-mygraph" .scale-to-fill}
Now think about the labels.

```r
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex)) +
  labs(
    x = "fastest speed ever driven (mph)",
    y = "grade-point average",
    title = "Not much overall relationship between speed and GPA!",
    subtitle = "(the fellers seem to drive faster and have lower GPAs)"
  )
```
:::




::: {focus-on="cr-mygraph" pan-to="0%, -90%" scale-by="2.0"}
Label for the x-axis:


```r
x = "fastest speed ever driven (mph)"
```

Include units (like "mph") whenever possible!
:::

@cr-mygraph

::: {focus-on="cr-mygraph" pan-to="90%, 0%" scale-by="2.0"}
Label for the y-axis:

```r
 y = "grade-point average"
```
::: 


@cr-mygraph

::: {focus-on="cr-mygraph" pan-to="30%, 90%" scale-by="1.7"}
Title and subtitle:

```r
title = "Not much overall relationship between speed and GPA!",
subtitle = "(the fellers seem to drive faster and have lower GPAs)"
```
**Pro Tip**:  The title should point out an important lesson conveyed by the plot.
:::




@cr-mygraph

:::{focus-on="cr-mygraph-caption" .scale-to-fill}
Working in Quarto?  Make a caption!


```{{r}}
#| fig-cap: "Scatterplot of GPA vs. fastest speed driven, for MAT 111 students."
ggplot(m111survey, aes(x = fastest, y = GPA)) +
  geom_point(aes(color = sex)) +
  labs(
    x = "fastest speed ever driven (mph)",
    y = "grade-point average",
    title = "Not much overall relationship between speed and GPA!",
    subtitle = "(the fellers seem to drive faster and have lower GPAs)"
  )
```

**Pro Tip**:  Unlike the title, the caption is neutral as to any information conveyed by the plot, but may provide additional context (in this case, the source of the data).

:::

:::

